---
title: "Calculating Fish Metrics"
author: "Karen Blocksom "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating Fish Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1 Preliminaries

One of the primary goals of this package is to calculate fish metrics as used in the National Rivers and Streams Assessment (NRSA) conducted in 2008-9 and 2013-14 by EPA's Office of Water. Fish were collected at all sites where present, and the common name, number of individuals, and number of anomalies were recorded in the field. Non-native status was then determined for all taxa on a site-specific basis. For the purposes of calculations, the metrics were broken down into several subsets, depending on which autecology traits were required to calculate them. In addition, if the non-native status is included with input data, the native version of metrics is also calculated, as well as alien species metrics, if desired. There are also separate functions to calculate just the metrics used in the regional multimetric indices (MMIs) and the MMI scores themselves based on metric inputs. This vignette will describe how to use the various functions along with the example datasets provided and the included taxa lists, to obtain the full set of metrics, a subset of metrics based on metric type, and the fish multimetric index used in assigning biological condition.

The functions in this package allow for a great deal of flexibility in terms of input variable names and the taxalist used in calculating metrics. However, the actual values of various traits, and the fish names as used in the taxalist are assumed to match those used in NRSA. This means that some metrics will not calculate correctly if the NRSA taxalist is not used. In addition, the functions assume that data have already been summarized to the level of sample and taxon, and a within-sample distinctness (0/1) already assigned to each taxon, so any steps required to get to this point must be done prior to use of the functions. All taxa in the count data must be included in the taxa list in order to be included in the calculations. Functions will look for specific types of inconsistencies, like missing variables, and either provide a warning or will fail to run, depending on the issue.  

********
# 2 Calculating NRSA fish metrics

There are two main approaches to calculating the fish metrics. One may either calculate all fish metrics at once, which may take a long time with larger datasets, or one may calculate various subsets of metrics. The latter approach may be helpful if only certain metrics are of interest or only certain traits are available. In the first approach, the function `calcAllFishMets` serves as a wrapper function for a number of metric subset functions. The other approach involves using individual funcions, including `calcFishTaxMets`, `calcFishTolMets`, `calcFishTrophicMets`, `calcFishOtherMets`, `calcFishNativeMets`, and/or `calcFishAnomMets`.

## 2.1 All metrics at once

Using the example plant data included in the package, `fishEx`, along with the included taxalist, `fishTaxa`.

```{r allFish.1}
library(aquametAlt)

head(fishEx)
```
This example dataset contains all of the necessary variables required to calculate all of the fish metrics. For this dataset, `UID` identifies individual samples, `TAXA_ID` identifies the taxon and links to same field in the taxalist, `FINAL_CT` is the count for the taxon, `ANOM_CT` is the number of anomalies for a given taxon, `IS_DISTINCT` is the distinctness, which 0 indicating non-distinct within the sample and 1 distinct, and `NON_NATIVE` indicates whether the taxon is non-native in this particular site (Y=YES, N=NO).

```{r allFish.2}
head(fishTaxa)
```
This is the desired format of the taxalist. All of the necessary traits are included in the taxalist for metric calculation. The variable `FAM_OR_CLS` is the same as `FAMILY` and is an artifact from the NRSA 2008-9 study.
```{r allFish.3}
outdf <- calcAllFishMets(fishEx,fishTaxa,sampID=c("UID"), dist="IS_DISTINCT",
                            ct="FINAL_CT",anomct='ANOM_CT',taxa_id='TAXA_ID',tol='TOLERANCE_NRSA'
                            , tolval='TOL_VAL_EMAPW', vel='VEL_NRSA', habitat='HABITAT_NRSA'
                            , trophic='TROPHIC_NRSA', migr='MIGR_NRSA', nonnat='NON_NATIVE'
                            , reprod='REPROD_NRSA', temp='TEMP_NRSA', family='FAMILY', genus='GENUS'
                            , comname='FINAL_NAME')

names(outdf)
```
The definitions of all variables can be found in **NRSA_Fish_Metric_Descriptions.pdf** found in this package.

## 2.2 Subsets of metrics



```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
